---
title: "Compressing multi allelic microhaplotypes"
format: html
editor: visual
---

## Compressing multi-allelic microhaplotypes to bi-allelic format

Building a genomic relatedness matrix (GRM) requires bi-allelic data. This is not compatible with our microhaplotype data, which sometimes has as many as 8 forms of one allele. Instead, we can compress this using a greedy algorithm. The greedy algorithm is a problem-solving algorithm that picks the most optimal choice at each step in the process. It does not consider previous choices- each choice is unique.

E. Anderson wrote the following code for A. Beulke for her PhD research. He gave her permission to share that file with me. I have adapted it a bit with additional comments to help my understanding and any replication of this process.

A concept he references frequently is minor allele frequencies (MAF's). We are shooting for MAF's of 0.5- this would show that the algorithm selected the most likely alleles at each step, thus working properly.

```{r}
library(tidyverse)
library(CKMRsim)
```

This is E. Anderson's code. First, he writes a function that acts any non-biallelic (SNP, or simple mhaps) into the algorithm.

```{r}
partition_greedy <- function(x) {
  
  if(length(x) == 2) {  # if it is already biallelic, just leave it
    return(
      tibble( # create a new data frame ewith tibble with the following col names
        AlleleOrig = names(x), # these are new column names we're creating for alleles that are already biallelic
        LumpedAllele = names(x),
        FreqOrig = unname(x) # strips out row and column labels
      )
    )
  } 
  # Sort descending to place large items first
  x_sorted <- sort(x, decreasing = TRUE)
  
  group1 <- c()
  group2 <- c()
  sum1 <- 0
  sum2 <- 0
  
  for (i in 1:length(x_sorted)) {
    val = x_sorted[i]
    if (sum1 <= sum2) {
      group1 <- c(group1, val)
      sum1 <- sum1 + unname(val)
    } else {
      group2 <- c(group2, val)
      sum2 <- sum2 + unname(val)
    }
  }
  
  # we use V and W to refer to the different lumped groups
  tibble(
    AlleleOrig = c(names(group1), names(group2)),
    LumpedAllele = c(rep("V", length(group1)), rep("W", length(group2))),
    FreqOrig = unname(c(group1, group2))
  )
}

```

Read in trimmed csv of all genetics data for Big Creek, trimmed of pseudoreplicates:

```{r}
bigcreek <- read.csv("Data/bigcreek_cleaned_snps_mhaps.csv")
nc <- ncol(bigcreek)
loci <- str_replace(names(bigcreek)[seq(2, nc, by = 2)], "\\.1$", "")
```

Now I'll read in my first .rds file. This file is just for Big Creek.

```{r}
bigcreek_freq = read_rds("Data/bigcreek_allelefreqs.rds")
str(bigcreek_freq$Allele)

lumpy <- bigcreek_freq %>%
  group_by(Chrom, Pos, Locus) %>%
  nest() %>%
  mutate(
    lumps = map(.x = data, .f = function(x) {
      v <- x$Freq
      names(v) <- x$Allele
      partition_greedy(v)
    })
  ) %>%
  unnest(cols = c(data, lumps)) %>%
  ungroup()

lumpy
```

I don't have AlleIdx or LocIdx, which Anne had. And I wonder if that's because my input data frame doesn't have the original genotypes in them. This wasn't a problem for the CKMRsim output, so I'll push forward, knowing this could be an issue later.

Now we look at how many alleles we end up with:

```{r}
new_freqs <- lumpy %>%
  group_by(Chrom, Pos, Locus, LumpedAllele) %>%
  summarise(
    LumpedFreq = sum(FreqOrig),
    orig_num_alle = n()
  )

new_freqs
```

At this point, my lumpedallele column is still using numbers, whereas his example has V and W. It still seemed to work, but this may be worth changing.

```{r}
MAFs <- new_freqs %>%
  filter(LumpedFreq == min(LumpedFreq)) %>%
  rename(MAF = LumpedFreq) %>%
  mutate(
    type = case_when(
      str_detect(Locus, "(hap1_code|haplotype)$") ~ "mhap",
      TRUE ~ "SNP"
    )
  )

ggplot(MAFs, aes(x = MAF, fill = type)) +
  geom_histogram(binwidth = 0.01) +
  facet_wrap(~ type, ncol = 1)
```

This didn't turn out nearly as pretty as E. Anderson's did on Anne's data. I'm not sure if something went wrong. I will proceed as if it's ok, but go over this in depth with Kerry once she returns from leave.

```{r}
joiner = lumpy %>% 
  dplyr::select(Chrom, Pos, Locus, AlleleOrig, LumpedAllele)
```

This data frame can then be joined with the long_genos data frame created in the the CKMR.Populations file. E. Anderson recommends using by = join_by(Chrom, Pos, Locus, Allele == AlleleOrig)

## Read in long genos from Big Creek

```{r}
lg <- read_rds("Data/bigcreek_longgenos.rds")
```

Join by locus? I don't fully understand how we're not losing data here.

```{r}
bigcreek_lumpy = left_join(
  lg, joiner, 
  by =  join_by(Locus, Allele == AlleleOrig)) %>% 
  dplyr::select(-Allele) %>%  # need to delete old allele column
  rename(Allele = LumpedAllele) # replace with new allele column
```

Now I'm going to go back through ckmrsim and see what happens.

Calculate allele frequencies:

```{r}
bigcreekalle_freqs <- bigcreek_lumpy %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))
```

```{r}
afreqs_ready <- reindex_markers(bigcreekalle_freqs)
```

Create a ckmr object

```{r}
bigcreek_ckmr_lumpy <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO","FS","U","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
bigcreek_ckmr_lumpy
```

Nice, so the greedy algorithm worked! The only concern is it says that it has between 2 and 2 alleles, and I think it should say between 1 and 2 alleles...

```{r}
bigcreek_Qs <- simulate_Qij(bigcreek_ckmr_lumpy, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
bigcreek_Qs
```

**Compute log-likelihood ratios to view them**

```{r}
bigcreek_FS_U_logls <- extract_logls(bigcreek_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

bigcreek_FS_U_logls
```

**Visualize**

```{r}
ggplot(bigcreek_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)
```

Check to make sure there are truly no identical individuals (there shouldn't be):

```{r}

bigcreek_lumpy_long_genos = bigcreek_lumpy %>% 
  dplyr::select(-c(Pos,Chrom))
matchers <- find_close_matching_genotypes(LG = bigcreek_lumpy_long_genos,
                                          CK = bigcreek_ckmr_lumpy,
                                          max_mismatch = 6)
matchers
```

Alright, so it worked!! Great! Let's see how many full sibs we have.

First, set loglr:

```{r}
bigcreek_FS_is <- mc_sample_simple(bigcreek_Qs, 
                 nu = "FS",
                 de = "U")

bigcreek_FS_is

#find where FNR > 0.01
bigcreek_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)

```

So logLR is 10. This makes some sense- we've collapsed some of the mhaps that gave us differentiation, so we lose some power.

```{r}
bigcreek_lumpy_pairwise_logls <- pairwise_kin_logl_ratios(
                                              D1 = bigcreek_lumpy_long_genos,
                                              D2 = bigcreek_lumpy_long_genos,
                                              CK = bigcreek_ckmr_lumpy,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)
```

Pairs with logLR greater than 10:

```{r}
bigcreeksibs<- bigcreek_lumpy_pairwise_logls %>%
  filter(logl_ratio > 10) %>%
  arrange(desc(logl_ratio))
```

We find 262 pairs. What about trios?

One of the next steps will be to build a sibling network and pull out groups. The goal should be to develop a workflow for each population that can be replicated for all other populations. During this network analysis, it will be important to consider year, as well.

# Pivot wider

I need to pivot the whole genome wider so that individuals stay as rows, but the loci and allele copies become columns.

```{r}
bigcreek_wider <- bigcreek_lumpy_long_genos %>% 
  pivot_wider(
    names_from = c(Locus, gene_copy),
    values_from = Allele,
    names_sep = "."
  )
```

To prepare it for AGH matrix, I now need to convert it into allele dosages. I will write this code with help of ChatGPT:

0 = zero copies of the reference allele 1 = one copy of the reference allele 2 = two copies of the reference allele NA = missing genotype

```{r}
library(dplyr)
library(stringr)

# Find the first allele-copy column for each locus, ending in .1
copy1_cols <- names(bigcreek_wider)[str_ends(names(bigcreek_wider), "\\.1$")]

# Start the dosage dataframe with individual ID (keep Indiv column)
bigcreek_dosage <- data.frame(
  Indiv = bigcreek_wider$Indiv
)

# Convert each pair of allele columns to dosage
for (col1 in copy1_cols) {
  
  # Find the matching second allele column
  col2 <- str_replace(col1, "\\.1$", ".2")
  
  # Get the locus name
  locus <- str_remove(col1, "\\.1$")
  
  # Find alleles present at this locus
  alleles <- sort(unique(c(
    bigcreek_wider[[col1]],
    bigcreek_wider[[col2]]
  )))
  
  # Remove NA from the list of alleles
  alleles <- alleles[!is.na(alleles)]
  
  # Check that the locus is biallelic
  if (length(alleles) > 2) {
    warning(
      paste(
        locus,
        "has", length(alleles),
        "alleles:", paste(alleles, collapse = ", ")
      )
    )
  }
  
  # Choose the first allele as the allele being counted
  ref_allele <- alleles[1]
  
  # Calculate dosage
  bigcreek_dosage[[locus]] <- ifelse(
    is.na(bigcreek_wider[[col1]]) |
      is.na(bigcreek_wider[[col2]]),
    NA,
    (bigcreek_wider[[col1]] == ref_allele) +
      (bigcreek_wider[[col2]] == ref_allele)
  )
}
```

```{r}
table(
  unlist(bigcreek_dosage[, -1]),
  useNA = "always"
)
```

I think this worked (but minimally checked ChatGPT's work on this, so not sure). I'll keep going and try the AGHmatrix code, which should be able to work directly with this.

# VanRaden matrix

```{r}
#Computing the additive relationship matrix based on VanRaden 2008
library(AGHmatrix)

# Make a copy so we don't alter the original data
bigcreek_G <- bigcreek_dosage
bigcreek_IDs <- bigcreek_dosage$Indiv

# Remove the individual ID column
bigcreek_G <- bigcreek_G[, -1]


# Convert the genotype data to numeric
bigcreek_G <- as.data.frame(
  lapply(bigcreek_G, as.numeric)
)


# Replace NA with -9
bigcreek_G[is.na(bigcreek_G)] <- -9


# Convert the data frame to a numeric matrix
bigcreek_G <- as.matrix(bigcreek_G)


# Check the structure
class(bigcreek_G)
mode(bigcreek_G)

G_VanRadenPine <- Gmatrix(
  SNPmatrix = bigcreek_G,
  missingValue = -9,
  maf = 0.05,
  method = "VanRaden"
)
```

```{r}
# ------------------------------------------------------------
# Add individual IDs to the genomic relationship matrix
# ------------------------------------------------------------

# Add individual IDs as row names
rownames(G_VanRadenPine) <- bigcreek_IDs

# Add individual IDs as column names
colnames(G_VanRadenPine) <- bigcreek_IDs


# ------------------------------------------------------------
# Look at the resulting relationship matrix
# ------------------------------------------------------------

G_VanRadenPine[1:10, 1:10]
```

```{r}
library(ggplot2)
library(reshape2)

# make the matrix into long format for plotting
longM <- melt(G_VanRadenPine)
max(longM$value)

ggplot(longM, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-2,2), space = "Lab") +
  theme_minimal() +
  theme(axis.text = element_text(angle = 45, hjust = 1, size = 2))+
  coord_fixed()

ggsave("BigCreek.Matrix.png", width = 5, height = 5, units = "in")
```

Let's make a histogram

```{r}
ggplot(longM, aes(x=value))+
  geom_histogram(binwidth = 0.05,
                 fill= "gray",
                 color = "black")+
  labs(x= "Relatedness",
       y = "",
       title = "Distribution of pairwise relatedness values in Big Creek Matrix")
```

Try inverting:

```{r}
G.inv <- solve(G_VanRadenPine)

# Add individual IDs as row names
rownames(G.inv) <- bigcreek_IDs

# Add individual IDs as column names
colnames(G.inv) <- bigcreek_IDs

# check the determinant.
det(G_VanRadenPine)
```

The determinant is very close to 0- not promising.

```{r}
eigen_G <- eigen(G_VanRadenPine, symmetric = TRUE)

eigen_G$values
min(eigen_G$values)

```

```{r}
max(eigen_G$values)
```

According to ChatGPT, the eigenvalues indicate a strong matrix for inversion. Yay! The next step will be to make a heat map of the matrix to make sure it worked.

```{r}
library(MCMCglmm)
```

```{r}
ctmax <- read.csv(here("Genomics/Data/Omy5_inversion_metadata_merge.csv"))
bigcreek_pheno <- ctmax %>% 
  filter(WATERSHED == "Big Creek") %>% 
  rename(Indiv = NMFS_DNA_ID_use) %>% 
  filter(!Indiv %in% c("M145524", "M145542"))

str(bigcreek_pheno$Treatment)
bigcreek_pheno$Treatment = as.character(bigcreek_pheno$Treatment)

class(G.inv)
# make the matrix into a sparse format
G.inv <- Matrix(G.inv, sparse = TRUE)

model1.2 <- MCMCglmm(LOE.Temp_C ~ 1 + Treatment, #Response and Fixed effect formula
                   random = ~Indiv, # Random effect formula
          ginverse = list(Indiv = G.inv), # correlations among random effect levels (here breeding values)
          data = bigcreek_pheno)
summary(model1.2)
```

Wow it worked! The next steps will be taking a slower walk through setting priors, checking how well the model ran, and interpreting the results.
