---
title: "Notebook 1: CKMRsim"
format: pdf
author: Paige Gardner, UC Santa Cruz pogardne@ucsc.edu
editor: visual
---

## CKMRsim: A Population level analysis

To do an initial assessment of relatedness, we use Eric Anderson's CKMRsim. Excellent tutorials on how to to use this pacakge can be found here: <https://eriqande.github.io/CKMRsim/articles/CKMRsim-example-1.html>

Here, we will assume that each population is isolated from one another. As a result, we will run this analysis on each population individually, but not separate by study year.

1.  Load in libraries and data
2.  Create new data frames for each watershed/population
3.  Run CKMR code on each population.

On 9/10, I re-ran all the code after filtering out loci noted by K Reid as either having low variability or probable correlation with other loci. I also used FNR = 0.01 as the threshold for setting logLR values for determining full-sibs instead of FPR.

## 1. Prep library and data

```{r}
#| output: false
library(tidyverse) 
library(MCMCpack) # for animal model
library(ggplot2) # plots
library(here) # easy data file access for Git
library(CKMRsim)
```

```{r}
#| output: false

# Phenotype and metadata
ctmax <- read.csv(here("Data/Omy5_inversion_metadata_merge.csv"))

# SNP data -------------------------------------------------------------------
# Data received 7/2/26
snp <- read.csv(here("Data/PaigeAdapCapFishSNPGenotypes.raw.csv"))
# extract first 7 characters into column with the same name as the ctmax dataframe
str(snp$Sample_ID)
snp$indiv.ID <- substr(snp$Sample_ID,1,7)
# remove original sample_ID column
snp <- snp %>% 
  dplyr::select(-Sample_ID)
snp <- snp %>% relocate(last_col())

### rename the columns ----------------------
# make the periods be _ for the next steps.
snp <- snp %>% 
  rename_with(~ gsub(".", "_",.x, fixed = TRUE))
# make note of the current order of loci in the data set
nc <- ncol(snp)
loci <- str_replace( # expects string, pattern, replacement
  names(snp)# returns the names of the columns
  [seq(2, nc, by = 2)], # creates a sequence of even numbers starting at 2 and ending at nc, stepping by 2
  "\\.\\.\\.[0-9]+$", "") # this cleans up column names. So I don't think this will impact my dataframe at all

#  reset the locus names
names(snp)[seq(2, nc, by = 2)] <- str_c(loci, "1", sep = ".") #put a .1 at the end of the first allele
names(snp)[seq(3, nc, by = 2)] <- str_c(loci, "2", sep = ".") # put a .2 at the end of the second allele
snp <- snp %>% 
  rename(indiv.ID = indiv_ID)


# MHap data----------------------------------------------------------------------
## Neutral
neutral_mhap <- read.csv(here("Data/Neutral_genotypes_mhapcode_wide.csv"))
nc <- ncol(neutral_mhap)
loci <- str_replace( # expects string, pattern, replacement
  names(neutral_mhap)# returns the names of the columns
  [seq(2, nc, by = 2)], # creates a sequence of even numbers starting at 2 and ending at nc, stepping by 2
  "\\.\\.\\.[0-9]+$", "") # this cleans up column names. So I don't think this will impact my dataframe at all

#  reset the locus names
names(neutral_mhap)[seq(2, nc, by = 2)] <- str_c(loci, "1", sep = ".") #put a .1 at the end of the first allele
names(neutral_mhap)[seq(3, nc, by = 2)] <- str_c(loci, "2", sep = ".") # put a .2 at the end of the second allele
```

### Important filtering from adaptive panel

We visually and graphically determined there was a lot of homogeneity in the adaptive panel that was skewing the relationships of individuals and making them look a lot more related than they are. Kerry suggested removing the following, across all panels:

-   SH127645_308: low variability

-   Omy_Ch22_VGLL3.6_snp2: low variability

-   Omy_Ch25_Six6_61317777: low variability

-   Omy_Ch25_Six6_61322413: low variability

-   Omy_Ch25_Six6.6_snp2: low variability

-   Omy_Ch28_11622532 : low variability

-   Omy_Ch28_11623412 : low variability

-   Omy_Ch28_11629294 : low variability

-   Omy_Ch28_11641623 : low variability

-   Omy_Ch28_11658853 : low variability

-   tag_id_2020 : low variability

-   tag_id_2192 : low variability

-   tag_id_3593 : low variability

-   tag_id_904 : low variability

-   SH114448_87: Potentially correlated

-   Omy_R04944 : Potentially correlated

-   SH107074_217 : Potentially correlated

-   Omy_Ch05_10_57417407 : Potentially correlated

-   Omy_Ch05_6_56087112: Potentially correlated

-   Omy_Ch05_7_76786553 : Potentially correlated

-   Omy_Ch05_8_28133607 : Potentially correlated

-   Omy_Ch05_9_54854574 : Potentially correlated

-   Omy_Ch25_Six6_61316270 : Potentially correlated

-   Omy_Ch25_Six6_61318852 : Potentially correlated

-   Omy_Ch25_Six6.6_snp1 : Potentially correlated

-   Omy_Ch28_11667578 : Potentially correlated

-   Omy_Ch28_11676622 : Potentially correlated

-   Omy_Ch28_11678603 : Potentially correlated

-   Omy_Ch28_11683310 : Potentially correlated

-   Omy_Ch28_11684744 : Potentially correlated

-   Omy_Ch28_greb1_mhap6 : Potentially correlated

-   Omy_Ch28_greb1_mhap8 : Potentially correlated

-   tag_id_3862 : Potentially correlated

-   tag_id_5746: Potentially correlated

```{r}
## Adaptive
adap_mhap <- read.csv(here("Data/Adaptive_genotypes_mhapcode_wide.csv"))

```

The ones Kerry suggested to remove, across all panels:

```{r}
merged <- merge(adap_mhap,neutral_mhap, by = "indiv.ID") 
fullmerge <- merge(snp, merged, by = "indiv.ID")
merged_clean <- fullmerge %>% 
  dplyr::select(!contains(c("SH127645_308",
                            "SexID",
                   "Omy_Ch22_VGLL3.6_snp2",
                   "Omy_Ch25_Six6_61317777",
                   "Omy_Ch25_Six6_61322413",
                   "Omy_Ch25_Six6.6_snp2",
                   "Omy_Ch28_11622532",
                   "Omy_Ch28_11623412",
                   "Omy_Ch28_11629294",
                   "Omy_Ch28_11641623",
                   "Omy_Ch28_11658853",
                   "tag_id_2020",
                   "tag_id_2192",
                   "tag_id_3593",
                   "tag_id_904",
                   "SH114448_87",
                   "Omy_R04944",
                   "SH107074_217",
                   "Omy_Ch05_10_57417407",
                   "Omy_Ch05_6_56087112",
                   "Omy_Ch05_7_76786553",
                   "Omy_Ch05_8_28133607",
                   "Omy_Ch05_9_54854574",
                   "Omy_Ch25_Six6_61316270",
                   "Omy_Ch25_Six6_61318852",
                   "Omy_Ch25_Six6.6_snp1",
                   "Omy_Ch28_11667578",
                   "Omy_Ch28_11676622",
                   "Omy_Ch28_11678603",
                   "Omy_Ch28_11683310",
                   "Omy_Ch28_11684744",
                   "Omy_Ch28_greb1_mhap6",
                   "Omy_Ch28_greb1_mhap8",
                   "tag_id_3862",
                   "tag_id_5746"))) %>% 
  filter(!indiv.ID %in% c("M145767","M145542", # matching indivs from scott creek and big creek
                       "M145524", "M145692"))
```

Lastly, all NA values should be 0's.

```{r}
sum(is.na(merged_clean))
```

Save the above data frame for easy access in the repository:

```{r}
write.csv(merged_clean, "Data/merged_TRIMMEDgenetics_KReid_9.10.26_codewide.csv", row.names = FALSE)
```

Merge location data from the metadata file with the above data file. (Most recently re-ran after filtering out loci suggested by K Reid on 9/10)

```{r}
populations <- ctmax %>% 
  rename(indiv.ID = NMFS_DNA_ID_use) %>% 
   filter(!indiv.ID %in% c("M145767","M145542", 
                       "M145524", "M145692",
                       "M145961" # this individual from RR doesn't have any genetic data
                       )) %>% 
  dplyr::select(c(indiv.ID,WATERSHED)) %>% 
  left_join(., merged_clean, by ="indiv.ID")

populations %>%
  count(indiv.ID) %>%
  filter(n > 1)
```

## 2. Separate Data Frames for Each Population

```{r}
bigcreek <- populations %>% 
  filter(WATERSHED == "Big Creek") %>% 
  dplyr::select(-WATERSHED)

write.csv(bigcreek, "Data/Populations/bigcreek_cleaned_snps_mhaps.csv")

scottcreek <- populations %>% 
  filter(WATERSHED == "Scott Creek")%>% 
  dplyr::select(-WATERSHED)

write.csv(scottcreek, "Data/Populations/scottcreek_cleaned_snps_mhaps.csv")

carmelriver <- populations %>% 
  filter(WATERSHED == "Carmel River")%>% 
  dplyr::select(-WATERSHED)

write.csv(carmelriver, "Data/Populations/carmelriver_cleaned_snps_mhaps.csv")

napariver <- populations %>% 
  filter(WATERSHED == "Napa River")%>% 
  dplyr::select(-WATERSHED)

write.csv(napariver, "Data/Populations/napariver_cleaned_snps_mhaps.csv")

russianriver <- populations %>% 
  filter(WATERSHED == "Russian River")%>% 
  dplyr::select(-WATERSHED)

write.csv(russianriver, "Data/Populations/russianriver_cleaned_snps_mhaps.csv")
```

## 3a) Big Sur

```{r}
# make note of the current order of loci in the data set
nc <- ncol(bigcreek)
loci <- str_replace(names(bigcreek)[seq(2, nc, by = 2)], "\\.1$", "")
# then make some long format genotypes
bigcreek_long_genos <- bigcreek %>% 
  gather(key = "loc", value = "Allele", -indiv.ID) %>%
   extract( # ChatGPT helped me with this code chunk. had to adapt from CKMR tutorial bc some of the gene names have other . in them
    loc,
    into = c("Locus", "gene_copy"),
    regex = "^(.*)\\.([12])$"
  ) %>%
  mutate(Allele = as.character(Allele)) %>%
  mutate(Allele = ifelse(Allele == "0", NA, Allele)) %>%
  rename(Indiv = indiv.ID)

saveRDS(bigcreek_long_genos, file = "Data/Populations/bigcreek_longgenos.rds")
```

Compute allele frequencies. Save it as an rds file for input into greedy algorithm (other script).

```{r}
bigcreekalle_freqs <- bigcreek_long_genos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))

# see what that looks like:
saveRDS(bigcreekalle_freqs, file = "Data/Populations/bigcreek_allelefreqs.rds")
```

Reindex the data allele frequency data frame so that ckmr can use it

```{r}
afreqs_ready <- reindex_markers(bigcreekalle_freqs)
```

**Create a ckmr object**

```{r}
bigcreek_ckmr <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO","FS","U","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
bigcreek_ckmr
```

**Simulate genotype pairs and calculate their log probabilities.**

```{r}
bigcreek_Qs <- simulate_Qij(bigcreek_ckmr, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
bigcreek_Qs
```

**Compute log-likelihood ratios to view them**

```{r}
bigcreek_FS_U_logls <- extract_logls(bigcreek_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

bigcreek_FS_U_logls
```

**Visualize**

```{r}
ggplot(bigcreek_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)
```

**Estimating FPR**

```{r}
bigcreek_FS_is <- mc_sample_simple(bigcreek_Qs, 
                 nu = "FS",
                 de = "U")

bigcreek_FS_is

FPR_results <- mc_sample_simple(
  bigcreek_Qs,
  nu = "FS",
  de = "U",
  lambda_stars = seq(5, 20, by = 0.1)
)
```

#### How many comparisons? Setting acceptable FPR

127 individuals

```{r}
0.1*((127*126)/2)^-1
```

## Setting logLR

Logl ratio that gives FNR = 0.01

```{r}
bigcreek_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)
```

We can set logLR to 18.

**Check for matchers.**

```{r}
matchers <- find_close_matching_genotypes(LG = long_genos,
                                          CK = bigcreek_ckmr,
                                          max_mismatch = 6)

matchers
```

None!

Compute LogLR for all pairwise comparisons.

```{r}
# Full sib pairs
bigcreek_fs_pairwise_logls <- pairwise_kin_logl_ratios(D1 = long_genos,
                                              D2 = long_genos,
                                              CK = bigcreek_ckmr,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)



# note, num_cores is set to 1 because more than that breaks
# it on CRAN checks.  But, leave it blank (i.e., don't use the
# num_core option) to, by default, use all
# your cores in parallel on a non-windows machine.
```

```{r}
fullsibs_bigcreek = bigcreek_fs_pairwise_logls %>%
  filter(logl_ratio > 18) %>%
  arrange(desc(logl_ratio))

bigcreek_FS_is %>%
  arrange(Lambda_star) %>%
  dplyr::select(Lambda_star, FPR) 
```

## 3b) Scott Creek

```{r}
nc <- ncol(scottcreek)
loci <- str_replace(names(scottcreek)[seq(2, nc, by = 2)], "\\.1$", "")
# then make some long format genotypes
scottcreek_long_genos <- scottcreek %>% 
  gather(key = "loc", value = "Allele", -indiv.ID) %>%
   extract( # ChatGPT helped me with this code chunk. had to adapt from CKMR tutorial bc some of the gene names have other . in them
    loc,
    into = c("Locus", "gene_copy"),
    regex = "^(.*)\\.([12])$"
  ) %>%
  mutate(Allele = as.character(Allele)) %>%
  mutate(Allele = ifelse(Allele == "0", NA, Allele)) %>%
  rename(Indiv = indiv.ID)

saveRDS(scottcreek_long_genos, file = "Data/Populations/scottcreek_longgenos.rds")
```

Compute allele frequencies.

```{r}
scottcreekalle_freqs <- scottcreek_long_genos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))

# see what that looks like:
scottcreekalle_freqs

saveRDS(scottcreekalle_freqs, file = "Data/Populations/scottcreek_allelefreqs.rds")
```

Reindex the data allele frequency data frame so that ckmr can use it

```{r}
afreqs_ready <- reindex_markers(scottcreekalle_freqs)
```

**Create a ckmr object**

```{r}
scottcreek_ckmr <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("FS", "U","PO","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
```

**Simulate genotype pairs and calculate their log probabilities.**

```{r}
scottcreek_Qs <- simulate_Qij(scottcreek_ckmr, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
scottcreek_Qs
```

**Compute log-likelihood ratios to view them**

```{r}
scottcreek_FS_U_logls <- extract_logls(scottcreek_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

scottcreek_FS_U_logls
```

**Visualize**

```{r}
ggplot(scottcreek_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)
```

### Estimating FPR

```{r}
scottcreek_FS_is <- mc_sample_simple(scottcreek_Qs, 
                 nu = "FS",
                 de = "U")

scottcreek_FS_is

FPR_results <- mc_sample_simple(
  scottcreek_Qs,
  nu = "FS",
  de = "U",
  tr = "U",
    lambda_stars = seq(0, 20, 1)
)
sum(is.na(scottcreek))
```

#### How many comparisons? Setting acceptable FPR

127 individuals

```{r}
0.1*((119*118)/2)^-1
```

#### Setting logLR

Logl ratio that gives FNR = 0.01

```{r}
scottcreek_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)
```

Set lambda to 17.

**Check for matchers.**

```{r}
matchers <- find_close_matching_genotypes(LG = long_genos,
                                          CK = scottcreek_ckmr,
                                          max_mismatch = 6)

matchers
```

Compute LogLR for all pairwise comparisons.

```{r}
# Full sib pairs
scottcreek_fs_pairwise_logls <- pairwise_kin_logl_ratios(D1 = long_genos,
                                              D2 = long_genos,
                                              CK = scottcreek_ckmr,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)



# note, num_cores is set to 1 because more than that breaks
# it on CRAN checks.  But, leave it blank (i.e., don't use the
# num_core option) to, by default, use all
# your cores in parallel on a non-windows machine.
```

```{r}
fullsibs_scottcreek<- scottcreek_fs_pairwise_logls %>%
  filter(logl_ratio > 16) %>%
  arrange(desc(logl_ratio))

scottcreek_FS_is %>%
  arrange(Lambda_star) %>%
  dplyr::select(Lambda_star, FPR) 
```

## 3c) Carmel River

```{r}
# make note of the current order of loci in the data set
nc <- ncol(carmelriver)
loci <- str_replace(names(carmelriver)[seq(2, nc, by = 2)], "\\.1$", "")
# then make some long format genotypes
carmelriver_longgenos <- carmelriver %>% 
  gather(key = "loc", value = "Allele", -indiv.ID) %>%
   extract( # ChatGPT helped me with this code chunk. had to adapt from CKMR tutorial bc some of the gene names have other . in them
    loc,
    into = c("Locus", "gene_copy"),
    regex = "^(.*)\\.([12])$"
  ) %>%
  mutate(Allele = as.character(Allele)) %>%
  mutate(Allele = ifelse(Allele == "0", NA, Allele)) %>%
  rename(Indiv = indiv.ID)

saveRDS(carmelriver_longgenos, file = "Data/Populations/carmelriver_longgenos.rds")
```

Compute allele frequencies.

```{r}
carmelriver_alle_freqs <- carmelriver_longgenos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))

saveRDS(carmelriver_alle_freqs, file = "Data/Populations/carmelriver_allelefreqs.rds")
# see what that looks like:
carmelriver_alle_freqs
```

Reindex the data allele frequency data frame so that ckmr can use it

```{r}
afreqs_ready <- reindex_markers(carmelriver_alle_freqs)
```

**Create a ckmr object**

```{r}
carmelriver_ckmr <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("FS", "U","PO","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
carmelriver_ckmr
```

**Simulate genotype pairs and calculate their log probabilities.**

```{r}
carmelriver_Qs <- simulate_Qij(carmelriver_ckmr, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
carmelriver_Qs
```

**Compute log-likelihood ratios to view them**

```{r}
carmelriver_FS_U_logls <- extract_logls(carmelriver_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

carmelriver_FS_U_logls
```

**Visualize**

```{r}
ggplot(carmelriver_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)
```

### Estimating FPR

```{r}
carmelriver_FS_is <- mc_sample_simple(carmelriver_Qs, 
                 nu = "FS",
                 de = "U")

carmelriver_FS_is

FPR_results <- mc_sample_simple(
  carmelriver_Qs,
  nu = "FS",
  de = "U",
  lambda_stars = seq(5, 20, by = 0.1)
)
```

#### Setting logLR

Logl ratio that gives FNR = 0.01

```{r}
carmelriver_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)
```

Set logLR to 15

**Check for matchers.**

```{r}
matchers <- find_close_matching_genotypes(LG = long_genos,
                                          CK = carmelriver_ckmr,
                                          max_mismatch = 6)

matchers
```

Compute LogLR for all pairwise comparisons.

```{r}
# Full sib pairs
carmelriver_fs_pairwise_logls <- pairwise_kin_logl_ratios(D1 = carmelriver_longgenos,
                                              D2 = carmelriver_longgenos,
                                              CK = carmelriver_ckmr,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)



# note, num_cores is set to 1 because more than that breaks
# it on CRAN checks.  But, leave it blank (i.e., don't use the
# num_core option) to, by default, use all
# your cores in parallel on a non-windows machine.
```

```{r}
fullsibs_carmelriver = carmelriver_fs_pairwise_logls %>%
  filter(logl_ratio > 15) %>%
  arrange(desc(logl_ratio))

carmelriver_FS_is %>%
  arrange(Lambda_star) %>%
  dplyr::select(Lambda_star, FPR) 
```

## 3d) Napa River

```{r}
# make note of the current order of loci in the data set
nc <- ncol(napariver)
loci <- str_replace(names(napariver)[seq(2, nc, by = 2)], "\\.1$", "")
# then make some long format genotypes
napariver_long_genos <- napariver %>% 
  gather(key = "loc", value = "Allele", -indiv.ID) %>%
   extract( # ChatGPT helped me with this code chunk. had to adapt from CKMR tutorial bc some of the gene names have other . in them
    loc,
    into = c("Locus", "gene_copy"),
    regex = "^(.*)\\.([12])$"
  ) %>%
  mutate(Allele = as.character(Allele)) %>%
  mutate(Allele = ifelse(Allele == "0", NA, Allele)) %>%
  rename(Indiv = indiv.ID)
saveRDS(napariver_long_genos, file = "Data/Populations/napariver_longgenos.rds")

```

Compute allele frequencies.

```{r}
napariver_alle_freqs <- napariver_long_genos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))

saveRDS(napariver_alle_freqs, file = "Data/Populations/napariver_allelefreqs.rds")

# see what that looks like:
napariver_alle_freqs
```

Reindex the data allele frequency data frame so that ckmr can use it

```{r}
afreqs_ready <- reindex_markers(napariver_alle_freqs)
```

**Create a ckmr object**

```{r}
napariver_ckmr <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("FS", "U","PO","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
napariver_ckmr
```

**Simulate genotype pairs and calculate their log probabilities.**

```{r}
napariver_Qs <- simulate_Qij(napariver_ckmr, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","PO","HS") )
napariver_Qs
```

**Compute log-likelihood ratios to view them**

```{r}
napariver_FS_U_logls <- extract_logls(napariver_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

napariver_FS_U_logls
```

**Visualize**

```{r}
ggplot(napariver_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)
```

### Estimating FPR

```{r}
napariver_FS_is <- mc_sample_simple(napariver_Qs, 
                 nu = "FS",
                 de = "U")

napariver_FS_is

FPR_results <- mc_sample_simple(
  napariver_Qs,
  nu = "FS",
  de = "U",
  lambda_stars = seq(5, 20, by = 0.1)
)
```

#### How many comparisons? Setting acceptable FPR

127 individuals

```{r}
0.1*((50*49)/2)^-1
```

#### Setting logLR

Logl ratio that gives FNR = 0.01

```{r}
napariver_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)
```

Set lambda to 12, or maybe 13.

**Check for matchers.**

```{r}
matchers <- find_close_matching_genotypes(LG = long_genos,
                                          CK = napariver_ckmr,
                                          max_mismatch = 6)

matchers
```

Compute LogLR for all pairwise comparisons.

```{r}
# Full sib pairs
napariver_fs_pairwise_logls <- pairwise_kin_logl_ratios(D1 = long_genos,
                                              D2 = long_genos,
                                              CK = napariver_ckmr,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)



# note, num_cores is set to 1 because more than that breaks
# it on CRAN checks.  But, leave it blank (i.e., don't use the
# num_core option) to, by default, use all
# your cores in parallel on a non-windows machine.
```

```{r}
fullsibs_napariver = napariver_fs_pairwise_logls %>%
  filter(logl_ratio > 13) %>%
  arrange(desc(logl_ratio))

napariver_FS_is %>%
  arrange(Lambda_star) %>%
  dplyr::select(Lambda_star, FPR) 
```

## 3e) Russian River

```{r}
# make note of the current order of loci in the data set
nc <- ncol(russianriver)
loci <- str_replace(names(russianriver)[seq(2, nc, by = 2)], "\\.1$", "")
# then make some long format genotypes
russianriver_long_genos <- russianriver %>% 
  gather(key = "loc", value = "Allele", -indiv.ID) %>%
   extract( # ChatGPT helped me with this code chunk. had to adapt from CKMR tutorial bc some of the gene names have other . in them
    loc,
    into = c("Locus", "gene_copy"),
    regex = "^(.*)\\.([12])$"
  ) %>%
  mutate(Allele = as.character(Allele)) %>%
  mutate(Allele = ifelse(Allele == "0", NA, Allele)) %>%
  rename(Indiv = indiv.ID)
saveRDS(russianriver_long_genos, file = "Data/Populations/russianriver_longgenos.rds")
```

Compute allele frequencies.

```{r}
russianriver_alle_freqs <- russianriver_long_genos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))
saveRDS(russianriver_alle_freqs, file = "Data/Populations/russianriver_allelefreqs.rds")
# see what that looks like:
russianriver_alle_freqs
```

Reindex the data allele frequency data frame so that ckmr can use it

```{r}
afreqs_ready <- reindex_markers(russianriver_alle_freqs)
```

**Create a ckmr object**

```{r}
russianriver_ckmr <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("FS", "U","PO","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
russianriver_ckmr
```

**Simulate genotype pairs and calculate their log probabilities.**

```{r}
russianriver_Qs <- simulate_Qij(russianriver_ckmr, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","PO","HS") )
russianriver_Qs
```

**Compute log-likelihood ratios to view them**

```{r}
russianriver_FS_U_logls <- extract_logls(russianriver_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

russianriver_FS_U_logls
```

**Visualize**

```{r}
ggplot(russianriver_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)
```

### Estimating FPR

```{r}
russianriver_FS_is <- mc_sample_simple(russianriver_Qs, 
                 nu = "FS",
                 de = "U")

russianriver_FS_is

FPR_results <- mc_sample_simple(
  russianriver_Qs,
  nu = "FS",
  de = "U",
  lambda_stars = seq(5, 20, by = 0.1)
)
```

#### How many comparisons? Setting acceptable FPR

127 individuals

```{r}
0.1*((72*71)/2)^-1
```

#### Setting logLR

Logl ratio that gives FNR = 0.01

```{r}
bigcreek_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)
```

Set lambda to either 17 or 18

**Check for matchers.**

```{r}
matchers <- find_close_matching_genotypes(LG = long_genos,
                                          CK = russianriver_ckmr,
                                          max_mismatch = 6)

matchers
```

Compute LogLR for all pairwise comparisons.

```{r}
# Full sib pairs
russianriver_fs_pairwise_logls <- pairwise_kin_logl_ratios(D1 = long_genos,
                                              D2 = long_genos,
                                              CK = russianriver_ckmr,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)



# note, num_cores is set to 1 because more than that breaks
# it on CRAN checks.  But, leave it blank (i.e., don't use the
# num_core option) to, by default, use all
# your cores in parallel on a non-windows machine.
```

```{r}
fullsibs_russianriver = russianriver_fs_pairwise_logls %>%
  filter(logl_ratio > 18) %>%
  arrange(desc(logl_ratio))

russianriver_FS_is %>%
  arrange(Lambda_star) %>%
  dplyr::select(Lambda_star, FPR) 
```

Whether or not I set the loglr to 17 or 18 doesn't change the number of full sibs.

## Check for monomorphic genes

```{r}
bigcreek %>% 
  dplyr::select(tag_id_2243_hap1_code.1)

length(unique(russianriver$tag_id_1637_hap1_code.2))
scottcreek %>%
  count(tag_id_986_hap1_code.2, sort = TRUE)
```
