---
title: "Notebook 3. Matrix"
format: html
editor: visual
---

# VanRaden matrix

```{r}
#Computing the additive relationship matrix based on VanRaden 2008
library(AGHmatrix)

# Make a copy so we don't alter the original data
bigcreek_G <- bigcreek_dosage
bigcreek_IDs <- bigcreek_dosage$Indiv

# Remove the individual ID column
bigcreek_G <- bigcreek_G[, -1]


# Convert the genotype data to numeric
bigcreek_G <- as.data.frame(
  lapply(bigcreek_G, as.numeric)
)


# Replace NA with -9
bigcreek_G[is.na(bigcreek_G)] <- -9


# Convert the data frame to a numeric matrix
bigcreek_G <- as.matrix(bigcreek_G)


# Check the structure
class(bigcreek_G)
mode(bigcreek_G)

G_VanRadenPine <- Gmatrix(
  SNPmatrix = bigcreek_G,
  missingValue = -9,
  maf = 0.05,
  method = "VanRaden"
)
```

```{r}
# ------------------------------------------------------------
# Add individual IDs to the genomic relationship matrix
# ------------------------------------------------------------

# Add individual IDs as row names
rownames(G_VanRadenPine) <- bigcreek_IDs

# Add individual IDs as column names
colnames(G_VanRadenPine) <- bigcreek_IDs


# ------------------------------------------------------------
# Look at the resulting relationship matrix
# ------------------------------------------------------------

G_VanRadenPine[1:10, 1:10]
```

```{r}
library(ggplot2)
library(reshape2)

# make the matrix into long format for plotting
longM <- melt(G_VanRadenPine)
max(longM$value)

ggplot(longM, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-2,2), space = "Lab") +
  theme_minimal() +
  theme(axis.text = element_text(angle = 45, hjust = 1, size = 2))+
  coord_fixed()

ggsave("BigCreek.Matrix.png", width = 5, height = 5, units = "in")
```

Let's make a histogram

```{r}
ggplot(longM, aes(x=value))+
  geom_histogram(binwidth = 0.05,
                 fill= "gray",
                 color = "black")+
  labs(x= "Relatedness",
       y = "",
       title = "Distribution of pairwise relatedness values in Big Creek Matrix")
```

Try inverting:

```{r}
G.inv <- solve(G_VanRadenPine)

# Add individual IDs as row names
rownames(G.inv) <- bigcreek_IDs

# Add individual IDs as column names
colnames(G.inv) <- bigcreek_IDs

# check the determinant.
det(G_VanRadenPine)
```

The determinant is very close to 0- not promising.

```{r}
eigen_G <- eigen(G_VanRadenPine, symmetric = TRUE)

eigen_G$values
min(eigen_G$values)

```

```{r}
max(eigen_G$values)
```

According to ChatGPT, the eigenvalues indicate a strong matrix for inversion. Yay! The next step will be to make a heat map of the matrix to make sure it worked.

```{r}
library(MCMCglmm)
```

```{r}
ctmax <- read.csv(here("Genomics/Data/Omy5_inversion_metadata_merge.csv"))
bigcreek_pheno <- ctmax %>% 
  filter(WATERSHED == "Big Creek") %>% 
  rename(Indiv = NMFS_DNA_ID_use) %>% 
  filter(!Indiv %in% c("M145524", "M145542"))

str(bigcreek_pheno$Treatment)
bigcreek_pheno$Treatment = as.character(bigcreek_pheno$Treatment)

class(G.inv)
# make the matrix into a sparse format
G.inv <- Matrix(G.inv, sparse = TRUE)

model1.2 <- MCMCglmm(LOE.Temp_C ~ 1 + Treatment, #Response and Fixed effect formula
                   random = ~Indiv, # Random effect formula
          ginverse = list(Indiv = G.inv), # correlations among random effect levels (here breeding values)
          data = bigcreek_pheno)
summary(model1.2)
```

Wow it worked! The next steps will be taking a slower walk through setting priors, checking how well the model ran, and interpreting the results.
