---
title: "Notebook 2: Compressing multi allelic microhaplotypes"
format: html
editor: visual
---

# Compressing multi-allelic microhaplotypes to bi-allelic format

Building a genomic relatedness matrix (GRM) requires bi-allelic data. This is not compatible with our microhaplotype data, which sometimes has as many as 8 forms of one allele. Instead, we can compress this using a greedy algorithm. The greedy algorithm is a problem-solving algorithm that picks the most optimal choice at each step in the process. It does not consider previous choices- each choice is unique.

E. Anderson wrote the following code for A. Beulke for her PhD research. He gave her permission to share that file with me. I have adapted it a bit with additional comments to help my understanding and any replication of this process.

A concept he references frequently is minor allele frequencies (MAF's). We are shooting for MAF's of 0.5- this would show that the algorithm selected the most likely alleles at each step, thus working properly.

```{r}
library(tidyverse)
library(CKMRsim)
library(dplyr)
library(stringr)
```

This is E. Anderson's code. First, he writes a function that acts any non-biallelic (SNP, or simple mhaps) into the algorithm. \# Algorithm

```{r}
partition_greedy <- function(x) {
  
  if(length(x) == 2) {  # if it is already biallelic, just leave it
    return(
      tibble( # create a new data frame ewith tibble with the following col names
        AlleleOrig = names(x), # these are new column names we're creating for alleles that are already biallelic
        LumpedAllele = names(x),
        FreqOrig = unname(x) # strips out row and column labels
      )
    )
  } 
  # Sort descending to place large items first
  x_sorted <- sort(x, decreasing = TRUE)
  
  group1 <- c()
  group2 <- c()
  sum1 <- 0
  sum2 <- 0
  
  for (i in 1:length(x_sorted)) {
    val = x_sorted[i]
    if (sum1 <= sum2) {
      group1 <- c(group1, val)
      sum1 <- sum1 + unname(val)
    } else {
      group2 <- c(group2, val)
      sum2 <- sum2 + unname(val)
    }
  }
  
  # we use V and W to refer to the different lumped groups
  tibble(
    AlleleOrig = c(names(group1), names(group2)),
    LumpedAllele = c(rep("V", length(group1)), rep("W", length(group2))),
    FreqOrig = unname(c(group1, group2))
  )
}
```

# 1) Big Creek

Read in trimmed csv of all genetics data for Big Creek, trimmed of pseudoreplicates:

```{r}
bigcreek <- read.csv("Data/Populations/bigcreek_cleaned_snps_mhaps.csv")
nc <- ncol(bigcreek)
loci <- str_replace(names(bigcreek)[seq(2, nc, by = 2)], "\\.1$", "")
```

Now I'll read in my first .rds file. This file is just for Big Creek.

```{r}
bigcreek_freq = read_rds("Data/Populations/bigcreek_allelefreqs.rds")
str(bigcreek_freq$Allele)

bc_lumpy <- bigcreek_freq %>%
  group_by(Chrom, Pos, Locus) %>%
  nest() %>%
  mutate(
    lumps = map(.x = data, .f = function(x) { # map applies a function to each element of a list and returns a list
      v <- x$Freq
      names(v) <- x$Allele
      partition_greedy(v)
    })
  ) %>%
  unnest(cols = c(data, lumps)) %>%
  ungroup()

bc_lumpy
```

I don't have AlleIdx or LocIdx. This wasn't a problem for the CKMRsim output, so I'll push forward, knowing this could be an issue later.

Now we look at how many alleles we end up with:

```{r}
bc_new_freqs <- bc_lumpy %>%
  group_by(Chrom, Pos, Locus, LumpedAllele) %>%
  summarise(
    LumpedFreq = sum(FreqOrig),
    orig_num_alle = n()
  )

bc_new_freqs
```

At this point, my lumpedallele column is still using numbers, whereas his example has V and W. It still seemed to work, but this may be worth changing.

```{r}
MAFs <- bc_new_freqs %>%
  filter(LumpedFreq == min(LumpedFreq)) %>%
  rename(MAF = LumpedFreq) %>%
  mutate(
    type = case_when(
      str_detect(Locus, "(hap1_code|haplotype)$") ~ "mhap",
      TRUE ~ "SNP"
    )
  )

ggplot(MAFs, aes(x = MAF, fill = type)) +
  geom_histogram(binwidth = 0.01, color="black") +
  facet_wrap(~ type, ncol = 1)+
  labs(y="",
       x= "MAF",
       title="Big Creek Greedy Alg. MAF")+
  theme_bw()
```

This didn't turn out nearly as pretty as E. Anderson's did on Anne's data. I'm not sure if something went wrong. I will proceed as if it's ok, but go over this in depth with Kerry once she returns from leave.

```{r}
joiner = lumpy %>% 
  dplyr::select(Chrom, Pos, Locus, AlleleOrig, LumpedAllele)
```

This data frame can then be joined with the long_genos data frame created in the the CKMR.Populations file. E. Anderson recommends using by = join_by(Chrom, Pos, Locus, Allele == AlleleOrig)

#### Read in long genos file

```{r}
lg <- read_rds("Data/Populations/bigcreek_longgenos.rds")
```

Join by locus? I don't fully understand how we're not losing data here.

```{r}
bigcreek_lumpy = left_join(
  lg, joiner, 
  by =  join_by(Locus, Allele == AlleleOrig)) %>% 
  dplyr::select(-Allele) %>%  # need to delete old allele column
  rename(Allele = LumpedAllele) # replace with new allele column
```

Now we run the new bi-allelic data into CKMRsim.

#### CKMRsim on newly biallelic data

Calculate allele frequencies:

```{r}
bigcreek_greedy_alle_freqs <- bigcreek_lumpy %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))
```

```{r}
afreqs_ready <- reindex_markers(bigcreek_greedy_alle_freqs)
```

Create a ckmr object

```{r}
bigcreek_ckmr_lumpy <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO","FS","U","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
bigcreek_ckmr_lumpy
```

Nice, so the greedy algorithm worked! The only concern is it says that it has between 2 and 2 alleles, and I think it should say between 1 and 2 alleles...

```{r}
bigcreek_Qs <- simulate_Qij(bigcreek_ckmr_lumpy, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
bigcreek_Qs
```

#### Visualize

```{r}
bigcreek_FS_U_logls <- extract_logls(bigcreek_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

bigcreek_FS_U_logls
```

```{r}
ggplot(bigcreek_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)+
  labs(title= "Big Creek Relatedness Likelihood, Biallelic")
```

Check to make sure there are truly no identical individuals (there shouldn't be):

```{r}
bigcreek_lumpy_long_genos = bigcreek_lumpy %>% 
  dplyr::select(-c(Pos,Chrom))
matchers <- find_close_matching_genotypes(LG = bigcreek_lumpy_long_genos,
                                          CK = bigcreek_ckmr_lumpy,
                                          max_mismatch = 6)
matchers
```

#### Check for full-sibs with bi-allelic data

```{r}
bigcreek_FS_is <- mc_sample_simple(bigcreek_Qs, 
                 nu = "FS",
                 de = "U")

bigcreek_FS_is

#find where FNR > 0.01
bigcreek_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)

```

So logLR is 10. This makes some sense- we've collapsed some of the mhaps that gave us differentiation, so we lose some power.

```{r}
bigcreek_lumpy_pairwise_logls <- pairwise_kin_logl_ratios(
                                              D1 = bigcreek_lumpy_long_genos,
                                              D2 = bigcreek_lumpy_long_genos,
                                              CK = bigcreek_ckmr_lumpy,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)
```

Pairs with logLR greater than 10:

```{r}
bigcreeksibs<- bigcreek_lumpy_pairwise_logls %>%
  filter(logl_ratio > 10) %>%
  arrange(desc(logl_ratio))
```

We find 262 pairs. What about trios?

One of the next steps will be to build a sibling network and pull out groups. The goal should be to develop a workflow for each population that can be replicated for all other populations. During this network analysis, it will be important to consider year, as well.

### Prep data for matrix

To use this data frame to construct a matrix, the whole genome data frame (converted to bi-allelic) needs to be pivoted wider so that individuals stay as rows, but the loci and allele copies become columns.

```{r}
bigcreek_wider <- bigcreek_lumpy_long_genos %>% 
  pivot_wider(
    names_from = c(Locus, gene_copy),
    values_from = Allele,
    names_sep = "."
  )
```

#### Allele dosages

To prepare it for AGH matrix, I now need to convert it into allele dosages. I wrote this code with the help of ChatGPT:

-   0 = zero copies of the reference allele

-   1 = one copy of the reference allele

-   2 = two copies of the reference allele

-   NA = missing genotype

```{r}
# Find the first allele-copy column for each locus, ending in .1
copy1_cols <- names(bigcreek_wider)[str_ends(names(bigcreek_wider), "\\.1$")]

# Start the dosage dataframe with individual ID (keep Indiv column)
bigcreek_dosage <- data.frame(
  Indiv = bigcreek_wider$Indiv
)

# Convert each pair of allele columns to dosage
for (col1 in copy1_cols) {
  
  # Find the matching second allele column
  col2 <- str_replace(col1, "\\.1$", ".2")
  
  # Get the locus name
  locus <- str_remove(col1, "\\.1$")
  
  # Find alleles present at this locus
  alleles <- sort(unique(c(
    bigcreek_wider[[col1]],
    bigcreek_wider[[col2]]
  )))
  
  # Remove NA from the list of alleles
  alleles <- alleles[!is.na(alleles)]
  
  # Check that the locus is biallelic
  if (length(alleles) > 2) {
    warning(
      paste(
        locus,
        "has", length(alleles),
        "alleles:", paste(alleles, collapse = ", ")
      )
    )
  }
  
  # Choose the first allele as the allele being counted
  ref_allele <- alleles[1]
  
  # Calculate dosage
  bigcreek_dosage[[locus]] <- ifelse(
    is.na(bigcreek_wider[[col1]]) |
      is.na(bigcreek_wider[[col2]]),
    NA,
    (bigcreek_wider[[col1]] == ref_allele) +
      (bigcreek_wider[[col2]] == ref_allele)
  )
}

write.csv(bigcreek_dosage, file = "Data/Populations/bigcreek_alleledosage.csv")
```

```{r}
#table(
 # unlist(bigcreek_dosage[, -1]),
#  useNA = "always"
#)
```

This appeared to work- we have half the number of observations as the full data frame, indicating one column per gene.

# 2) Scott Creek

Follow the same steps as above.

```{r}
scottcreek <- read.csv("Data/Populations/scottcreek_cleaned_snps_mhaps.csv")
nc <- ncol(scottcreek)
loci <- str_replace(names(scottcreek)[seq(2, nc, by = 2)], "\\.1$", "")
```

Read in .rds file for Scott Creek.

```{r}
scottcreek_freq = read_rds("Data/Populations/scottcreek_allelefreqs.rds")
str(scottcreek_freq$Allele)

sc_lumpy <- scottcreek_freq %>%
  group_by(Chrom, Pos, Locus) %>%
  nest() %>%
  mutate(
    lumps = map(.x = data, .f = function(x) { # map applies a function to each element of a list and returns a list
      v <- x$Freq
      names(v) <- x$Allele
      partition_greedy(v)
    })
  ) %>%
  unnest(cols = c(data, lumps)) %>%
  ungroup()

sc_lumpy
```

```{r}
sc_new_freqs <- sc_lumpy %>%
  group_by(Chrom, Pos, Locus, LumpedAllele) %>%
  summarise(
    LumpedFreq = sum(FreqOrig),
    orig_num_alle = n()
  )

sc_new_freqs
```

Visualize.

```{r}
MAFs <- sc_new_freqs %>%
  filter(LumpedFreq == min(LumpedFreq)) %>%
  rename(MAF = LumpedFreq) %>%
  mutate(
    type = case_when(
      str_detect(Locus, "(hap1_code|haplotype)$") ~ "mhap",
      TRUE ~ "SNP"
    )
  )

ggplot(MAFs, aes(x = MAF, fill = type)) +
  geom_histogram(binwidth = 0.01, color="black") +
  facet_wrap(~ type, ncol = 1)+
  labs(y="",
       x= "MAF",
       title="Scott Creek Greedy Alg. MAF")+
  theme_bw()
MAFs %>% 
  filter(MAF > 0.9)

scottcreek 
```

```{r}
joiner = sc_lumpy %>% 
  dplyr::select(Chrom, Pos, Locus, AlleleOrig, LumpedAllele)
```

This data frame can then be joined with the long_genos data frame created in the the CKMR.Populations file. E. Anderson recommends using by = join_by(Chrom, Pos, Locus, Allele == AlleleOrig)

#### Read in long genos file

```{r}
sc_lg <- read_rds("Data/Populations/scottcreek_longgenos.rds")
```

Join by locus.

```{r}
scottcreek_lumpy = left_join(
  sc_lg, joiner, 
  by =  join_by(Locus, Allele == AlleleOrig)) %>% 
  dplyr::select(-Allele) %>%  # need to delete old allele column
  rename(Allele = LumpedAllele) # replace with new allele column
```

Now we run the new bi-allelic data into CKMRsim.

#### CKMRsim on newly biallelic data

Calculate allele frequencies:

```{r}
scottcreek_greedy_alle_freqs <- scottcreek_lumpy %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))
```

```{r}
afreqs_ready <- reindex_markers(scottcreek_greedy_alle_freqs)
```

Create a ckmr object

```{r}
scottcreek_ckmr_lumpy <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO","FS","U","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
scottcreek_ckmr_lumpy
```

```{r}
scottcreek_Qs <- simulate_Qij(scottcreek_ckmr_lumpy, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
scottcreek_Qs
```

#### Visualize

```{r}
scottcreek_FS_U_logls <- extract_logls(scottcreek_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

scottcreek_FS_U_logls
```

```{r}
ggplot(scottcreek_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)+
  labs(title= "Scott Creek Relatedness Likelihood, Biallelic")
```

Check to make sure there are truly no identical individuals (there shouldn't be):

```{r}
scottcreek_lumpy_long_genos = scottcreek_lumpy %>% 
  dplyr::select(-c(Pos,Chrom))
matchers <- find_close_matching_genotypes(LG = scottcreek_lumpy_long_genos,
                                          CK = scottcreek_ckmr_lumpy,
                                          max_mismatch = 6)
matchers
```

No matchers found.

#### Check for full-sibs with bi-allelic data

```{r}
scottcreek_FS_is <- mc_sample_simple(scottcreek_Qs, 
                 nu = "FS",
                 de = "U")

scottcreek_FS_is

#find where FNR > 0.01
scottcreek_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)

```

Using a FNR threshold of 0.01, lambda should be set to 9 or 10 (calculated value: 9.45).

```{r}
scottcreek_lumpy_pairwise_logls <- pairwise_kin_logl_ratios(
                                              D1 = scottcreek_lumpy_long_genos,
                                              D2 = scottcreek_lumpy_long_genos,
                                              CK = scottcreek_ckmr_lumpy,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)
```

Pairs with logLR greater than 10:

```{r}
scottcreeksibs<- scottcreek_lumpy_pairwise_logls %>%
  filter(logl_ratio > 9) %>%
  arrange(desc(logl_ratio))
```

We find 271 pairs.

### Prep data for matrix

To use this data frame to construct a matrix, the whole genome data frame (converted to bi-allelic) needs to be pivoted wider so that individuals stay as rows, but the loci and allele copies become columns.

```{r}
scottcreek_wider <- scottcreek_lumpy_long_genos %>% 
  pivot_wider(
    names_from = c(Locus, gene_copy),
    values_from = Allele,
    names_sep = "."
  )
```

#### Allele dosages

To prepare it for AGH matrix, I now need to convert it into allele dosages. I wrote this code with the help of ChatGPT:

-   0 = zero copies of the reference allele

-   1 = one copy of the reference allele

-   2 = two copies of the reference allele

-   NA = missing genotype

```{r}
# Find the first allele-copy column for each locus, ending in .1
copy1_cols <- names(scottcreek_wider)[str_ends(names(scottcreek_wider), "\\.1$")]

# Start the dosage dataframe with individual ID (keep Indiv column)
scottcreek_dosage <- data.frame(
  Indiv = scottcreek_wider$Indiv
)

# Convert each pair of allele columns to dosage
for (col1 in copy1_cols) {
  
  # Find the matching second allele column
  col2 <- str_replace(col1, "\\.1$", ".2")
  
  # Get the locus name
  locus <- str_remove(col1, "\\.1$")
  
  # Find alleles present at this locus
  alleles <- sort(unique(c(
    scottcreek_wider[[col1]],
    scottcreek_wider[[col2]]
  )))
  
  # Remove NA from the list of alleles
  alleles <- alleles[!is.na(alleles)]
  
  # Check that the locus is biallelic
  if (length(alleles) > 2) {
    warning(
      paste(
        locus,
        "has", length(alleles),
        "alleles:", paste(alleles, collapse = ", ")
      )
    )
  }
  
  # Choose the first allele as the allele being counted
  ref_allele <- alleles[1]
  
  # Calculate dosage
  scottcreek_dosage[[locus]] <- ifelse(
    is.na(scottcreek_wider[[col1]]) |
      is.na(scottcreek_wider[[col2]]),
    NA,
    (scottcreek_wider[[col1]] == ref_allele) +
      (scottcreek_wider[[col2]] == ref_allele)
  )
}

write.csv(scottcreek_dosage, file = "Data/Populations/scottcreek_alleledosage.csv")
```

# 3) Carmel River

Follow the same steps as above.

```{r}
carmelriver <- read.csv("Data/Populations/carmelriver_cleaned_snps_mhaps.csv")
nc <- ncol(carmelriver)
loci <- str_replace(names(carmelriver)[seq(2, nc, by = 2)], "\\.1$", "")
```

Now I'll read in my first .rds file. This file is just for Big Creek.

```{r}
carmelriver_freq = read_rds("Data/Populations/carmelriver_allelefreqs.rds")
str(carmelriver_freq$Allele)
# Run the algorithm
cr_lumpy <- carmelriver_freq %>%
  group_by(Chrom, Pos, Locus) %>%
  nest() %>%
  mutate(
    lumps = map(.x = data, .f = function(x) { # map applies a function to each element of a list and returns a list
      v <- x$Freq
      names(v) <- x$Allele
      partition_greedy(v)
    })
  ) %>%
  unnest(cols = c(data, lumps)) %>%
  ungroup()

cr_lumpy
```

```{r}
cr_new_freqs <- cr_lumpy %>%
  group_by(Chrom, Pos, Locus, LumpedAllele) %>%
  summarise(
    LumpedFreq = sum(FreqOrig),
    orig_num_alle = n()
  )

cr_new_freqs
```

Visualize.

```{r}
MAFs <- cr_new_freqs %>%
  filter(LumpedFreq == min(LumpedFreq)) %>%
  rename(MAF = LumpedFreq) %>%
  mutate(
    type = case_when(
      str_detect(Locus, "(hap1_code|haplotype)$") ~ "mhap",
      TRUE ~ "SNP"
    )
  )

ggplot(MAFs, aes(x = MAF, fill = type)) +
  geom_histogram(binwidth = 0.01, color="black") +
  facet_wrap(~ type, ncol = 1)+
  labs(y="",
       x= "MAF",
       title="Carmel River Greedy Alg. MAF")+
  theme_bw()
```

```{r}
joiner = cr_lumpy %>% 
  dplyr::select(Chrom, Pos, Locus, AlleleOrig, LumpedAllele)
```

This data frame can then be joined with the long_genos data frame created in the the CKMR.Populations file. E. Anderson recommends using by = join_by(Chrom, Pos, Locus, Allele == AlleleOrig)

#### Read in long genos file

```{r}
lg <- read_rds("Data/Populations/carmelriver_longgenos.rds")
```

Join by locus? I don't fully understand how we're not losing data here.

```{r}
carmelriver_lumpy = left_join(
  lg, joiner, 
  by =  join_by(Locus, Allele == AlleleOrig)) %>% 
  dplyr::select(-Allele) %>%  # need to delete old allele column
  rename(Allele = LumpedAllele) # replace with new allele column
```

Now we run the new bi-allelic data into CKMRsim.

#### CKMRsim on newly biallelic data

Calculate allele frequencies:

```{r}
carmelriver_greedy_alle_freqs <- carmelriver_lumpy %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))
```

```{r}
afreqs_ready <- reindex_markers(carmelriver_greedy_alle_freqs)
```

Create a ckmr object

```{r}
carmelriver_ckmr_lumpy <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO","FS","U","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
carmelriver_ckmr_lumpy
```

```{r}
carmelriver_Qs <- simulate_Qij(carmelriver_ckmr_lumpy, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
carmelriver_Qs
```

#### Visualize

```{r}
carmelriver_FS_U_logls <- extract_logls(carmelriver_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

carmelriver_FS_U_logls
```

```{r}
ggplot(carmelriver_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)+
  labs(title= "Big Creek Relatedness Likelihood, Biallelic")
```

Check to make sure there are truly no identical individuals (there shouldn't be):

```{r}
carmelriver_lumpy_long_genos = carmelriver_lumpy %>% 
  dplyr::select(-c(Pos,Chrom))
matchers <- find_close_matching_genotypes(LG = carmelriver_lumpy_long_genos,
                                          CK = carmelriver_ckmr_lumpy,
                                          max_mismatch = 6)
matchers
```

#### Check for full-sibs with bi-allelic data

```{r}
carmelriver_FS_is <- mc_sample_simple(carmelriver_Qs, 
                 nu = "FS",
                 de = "U")

carmelriver_FS_is

#find where FNR > 0.01
carmelriver_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)

```

I will set the lambda to 11.

```{r}
carmelriver_lumpy_pairwise_logls <- pairwise_kin_logl_ratios(
                                              D1 = carmelriver_lumpy_long_genos,
                                              D2 = carmelriver_lumpy_long_genos,
                                              CK = carmelriver_ckmr_lumpy,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)
```

Pairs with logLR greater than 11:

```{r}
carmelriversibs<- carmelriver_lumpy_pairwise_logls %>%
  filter(logl_ratio > 11) %>%
  arrange(desc(logl_ratio))
```

We find 22 pairs.Before the biallelic conversion, we found 23.

### Prep data for matrix

To use this data frame to construct a matrix, the whole genome data frame (converted to bi-allelic) needs to be pivoted wider so that individuals stay as rows, but the loci and allele copies become columns.

```{r}
carmelriver_wider <- carmelriver_lumpy_long_genos %>% 
  pivot_wider(
    names_from = c(Locus, gene_copy),
    values_from = Allele,
    names_sep = "."
  )
```

#### Allele dosages

To prepare it for AGH matrix, I now need to convert it into allele dosages. I wrote this code with the help of ChatGPT:

-   0 = zero copies of the reference allele

-   1 = one copy of the reference allele

-   2 = two copies of the reference allele

-   NA = missing genotype

```{r}
# Find the first allele-copy column for each locus, ending in .1
copy1_cols <- names(carmelriver_wider)[str_ends(names(carmelriver_wider), "\\.1$")]

# Start the dosage dataframe with individual ID (keep Indiv column)
carmelriver_dosage <- data.frame(
  Indiv = carmelriver_wider$Indiv
)

# Convert each pair of allele columns to dosage
for (col1 in copy1_cols) {
  
  # Find the matching second allele column
  col2 <- str_replace(col1, "\\.1$", ".2")
  
  # Get the locus name
  locus <- str_remove(col1, "\\.1$")
  
  # Find alleles present at this locus
  alleles <- sort(unique(c(
    carmelriver_wider[[col1]],
    carmelriver_wider[[col2]]
  )))
  
  # Remove NA from the list of alleles
  alleles <- alleles[!is.na(alleles)]
  
  # Check that the locus is biallelic
  if (length(alleles) > 2) {
    warning(
      paste(
        locus,
        "has", length(alleles),
        "alleles:", paste(alleles, collapse = ", ")
      )
    )
  }
  
  # Choose the first allele as the allele being counted
  ref_allele <- alleles[1]
  
  # Calculate dosage
  carmelriver_dosage[[locus]] <- ifelse(
    is.na(carmelriver_wider[[col1]]) |
      is.na(carmelriver_wider[[col2]]),
    NA,
    (carmelriver_wider[[col1]] == ref_allele) +
      (carmelriver_wider[[col2]] == ref_allele)
  )
}

write.csv(carmelriver_dosage, file = "Data/Populations/carmelriver_alleledosage.csv")
```

# 4) Napa River

Follow the same steps as above.

```{r}
napariver <- read.csv("Data/Populations/napariver_cleaned_snps_mhaps.csv")
nc <- ncol(napariver)
loci <- str_replace(names(napariver)[seq(2, nc, by = 2)], "\\.1$", "")
```

Now I'll read in my first .rds file. This file is just for Big Creek.

```{r}
napariver_freq = read_rds("Data/Populations/napariver_allelefreqs.rds")
str(napariver_freq$Allele)

nr_lumpy <- napariver_freq %>%
  group_by(Chrom, Pos, Locus) %>%
  nest() %>%
  mutate(
    lumps = map(.x = data, .f = function(x) { # map applies a function to each element of a list and returns a list
      v <- x$Freq
      names(v) <- x$Allele
      partition_greedy(v)
    })
  ) %>%
  unnest(cols = c(data, lumps)) %>%
  ungroup()

nr_lumpy
```

```{r}
nr_new_freqs <- nr_lumpy %>%
  group_by(Chrom, Pos, Locus, LumpedAllele) %>%
  summarise(
    LumpedFreq = sum(FreqOrig),
    orig_num_alle = n()
  )

nr_new_freqs
```

Visualize.

```{r}
MAFs <- nr_new_freqs %>%
  filter(LumpedFreq == min(LumpedFreq)) %>%
  rename(MAF = LumpedFreq) %>%
  mutate(
    type = case_when(
      str_detect(Locus, "(hap1_code|haplotype)$") ~ "mhap",
      TRUE ~ "SNP"
    )
  )

ggplot(MAFs, aes(x = MAF, fill = type)) +
  geom_histogram(binwidth = 0.01, color="black") +
  facet_wrap(~ type, ncol = 1)+
  labs(y="",
       x= "MAF",
       title="Napa River Greedy Alg. MAF")+
  theme_bw()

MAFs %>% 
  filter(MAF >0.9)

napariver %>%
  count(tag_id_4933_hap1_code.2, sort = TRUE)
```

```{r}
joiner = nr_lumpy %>% 
  dplyr::select(Chrom, Pos, Locus, AlleleOrig, LumpedAllele)
```

This data frame can then be joined with the long_genos data frame created in the the CKMR.Populations file. E. Anderson recommends using by = join_by(Chrom, Pos, Locus, Allele == AlleleOrig)

#### Read in long genos file

```{r}
lg <- read_rds("Data/Populations/napariver_longgenos.rds")
```

Join by locus.

```{r}
napariver_lumpy = left_join(
  lg, joiner, 
  by =  join_by(Locus, Allele == AlleleOrig)) %>% 
  dplyr::select(-Allele) %>%  # need to delete old allele column
  rename(Allele = LumpedAllele) # replace with new allele column
```

Now we run the new bi-allelic data into CKMRsim.

#### CKMRsim on newly biallelic data

Calculate allele frequencies:

```{r}
napariver_greedy_alle_freqs <- napariver_lumpy %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))
```

```{r}
afreqs_ready <- reindex_markers(napariver_greedy_alle_freqs)
```

Create a ckmr object

```{r}
napariver_ckmr_lumpy <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO","FS","U","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
napariver_ckmr_lumpy
```

```{r}
napariver_Qs <- simulate_Qij(napariver_ckmr_lumpy, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
napariver_Qs
```

#### Visualize

```{r}
napariver_FS_U_logls <- extract_logls(napariver_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

napariver_FS_U_logls
```

```{r}
ggplot(napariver_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)+
  labs(title= "Napa River Relatedness Likelihood, Biallelic")
```

Check to make sure there are truly no identical individuals (there shouldn't be):

```{r}
napariver_lumpy_long_genos = napariver_lumpy %>% 
  dplyr::select(-c(Pos,Chrom))
matchers <- find_close_matching_genotypes(LG = napariver_lumpy_long_genos,
                                          CK = napariver_ckmr_lumpy,
                                          max_mismatch = 6)
matchers
```

#### Check for full-sibs with bi-allelic data

```{r}
napariver_FS_is <- mc_sample_simple(napariver_Qs, 
                 nu = "FS",
                 de = "U")

napariver_FS_is

#find where FNR > 0.01
napariver_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)

```

Set lambda to 9.

```{r}
napariver_lumpy_pairwise_logls <- pairwise_kin_logl_ratios(
                                              D1 = napariver_lumpy_long_genos,
                                              D2 = napariver_lumpy_long_genos,
                                              CK = napariver_ckmr_lumpy,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)
```

Pairs with logLR greater than 9:

```{r}
napariversibs<- napariver_lumpy_pairwise_logls %>%
  filter(logl_ratio > 9) %>%
  arrange(desc(logl_ratio))
```

We find 83 pairs.

### Prep data for matrix

To use this data frame to construct a matrix, the whole genome data frame (converted to bi-allelic) needs to be pivoted wider so that individuals stay as rows, but the loci and allele copies become columns.

```{r}
napariver_wider <- napariver_lumpy_long_genos %>% 
  pivot_wider(
    names_from = c(Locus, gene_copy),
    values_from = Allele,
    names_sep = "."
  )
```

#### Allele dosages

To prepare it for AGH matrix, I now need to convert it into allele dosages. I wrote this code with the help of ChatGPT:

-   0 = zero copies of the reference allele

-   1 = one copy of the reference allele

-   2 = two copies of the reference allele

-   NA = missing genotype

```{r}
# Find the first allele-copy column for each locus, ending in .1
copy1_cols <- names(napariver_wider)[str_ends(names(napariver_wider), "\\.1$")]

# Start the dosage dataframe with individual ID (keep Indiv column)
napariver_dosage <- data.frame(
  Indiv = napariver_wider$Indiv
)

# Convert each pair of allele columns to dosage
for (col1 in copy1_cols) {
  
  # Find the matching second allele column
  col2 <- str_replace(col1, "\\.1$", ".2")
  
  # Get the locus name
  locus <- str_remove(col1, "\\.1$")
  
  # Find alleles present at this locus
  alleles <- sort(unique(c(
    napariver_wider[[col1]],
    napariver_wider[[col2]]
  )))
  
  # Remove NA from the list of alleles
  alleles <- alleles[!is.na(alleles)]
  
  # Check that the locus is biallelic
  if (length(alleles) > 2) {
    warning(
      paste(
        locus,
        "has", length(alleles),
        "alleles:", paste(alleles, collapse = ", ")
      )
    )
  }
  
  # Choose the first allele as the allele being counted
  ref_allele <- alleles[1]
  
  # Calculate dosage
  napariver_dosage[[locus]] <- ifelse(
    is.na(napariver_wider[[col1]]) |
      is.na(napariver_wider[[col2]]),
    NA,
    (napariver_wider[[col1]] == ref_allele) +
      (napariver_wider[[col2]] == ref_allele)
  )
}

write.csv(napariver_dosage, file = "Data/Populations/napariver_alleledosage.csv")
```

# 5) Russian River

Follow the same steps as above.

```{r}
russianriver <- read.csv("Data/Populations/russianriver_cleaned_snps_mhaps.csv")
nc <- ncol(russianriver)
loci <- str_replace(names(russianriver)[seq(2, nc, by = 2)], "\\.1$", "")
sum(is.na(russianriver)) # double check no NA values
```

Read in allele frequency rds file.

```{r}
russianriver_freq = read_rds("Data/Populations/russianriver_allelefreqs.rds")
str(russianriver_freq$Allele)

rr_lumpy <- russianriver_freq %>%
  group_by(Chrom, Pos, Locus) %>%
  nest() %>%
  mutate(
    lumps = map(.x = data, .f = function(x) { # map applies a function to each element of a list and returns a list
      v <- x$Freq
      names(v) <- x$Allele
      partition_greedy(v)
    })
  ) %>%
  unnest(cols = c(data, lumps)) %>%
  ungroup()

rr_lumpy
```

```{r}
rr_new_freqs <- rr_lumpy %>%
  group_by(Chrom, Pos, Locus, LumpedAllele) %>%
  summarise(
    LumpedFreq = sum(FreqOrig),
    orig_num_alle = n()
  )

rr_new_freqs
```

Visualize.

```{r}
MAFs <- rr_new_freqs %>%
  filter(LumpedFreq == min(LumpedFreq)) %>%
  rename(MAF = LumpedFreq) %>%
  mutate(
    type = case_when(
      str_detect(Locus, "(hap1_code|haplotype)$") ~ "mhap",
      TRUE ~ "SNP"
    )
  )


ggplot(MAFs, aes(x = MAF, fill = type)) +
  geom_histogram(binwidth = 0.01, color="black") +
  facet_wrap(~ type, ncol = 1)+
  labs(y="",
       x= "MAF",
       title="Russian River Greedy Alg. MAF")+
  theme_bw()
```

```{r}
joiner = rr_lumpy %>% 
  dplyr::select(Chrom, Pos, Locus, AlleleOrig, LumpedAllele)
```

This data frame can then be joined with the long_genos data frame created in the the CKMR.Populations file. E. Anderson recommends using by = join_by(Chrom, Pos, Locus, Allele == AlleleOrig)

#### Read in long genos file

```{r}
lg <- read_rds("Data/Populations/russianriver_longgenos.rds")
```

Join by locus? I don't fully understand how we're not losing data here.

```{r}
russianriver_lumpy = left_join(
  lg, joiner, 
  by =  join_by(Locus, Allele == AlleleOrig)) %>% 
  dplyr::select(-Allele) %>%  # need to delete old allele column
  rename(Allele = LumpedAllele) # replace with new allele column
```

Now we run the new bi-allelic data into CKMRsim.

#### CKMRsim on newly biallelic data

Calculate allele frequencies:

```{r}
russianriver_greedy_alle_freqs <- russianriver_lumpy %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  dplyr::select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele))
```

```{r}
afreqs_ready <- reindex_markers(russianriver_greedy_alle_freqs)
```

Create a ckmr object

```{r}
russianriver_ckmr_lumpy <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO","FS","U","HS"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
russianriver_ckmr_lumpy
```

```{r}
russianriver_Qs <- simulate_Qij(russianriver_ckmr_lumpy, 
                       calc_relats = c("FS", "U","PO"),
                       sim_relats = c("FS",  "U","HS","PO") )
russianriver_Qs
```

#### Visualize

```{r}
russianriver_FS_U_logls <- extract_logls(russianriver_Qs,
                            numer = c(FS = 1),
                            denom = c(U = 1))

russianriver_FS_U_logls
```

```{r}
ggplot(russianriver_FS_U_logls,
            aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25)+
  labs(title= "Russian River Relatedness Likelihood, Biallelic")
```

Check to make sure there are truly no identical individuals (there shouldn't be):

```{r}
russianriver_lumpy_long_genos = russianriver_lumpy %>% 
  dplyr::select(-c(Pos,Chrom))
matchers <- find_close_matching_genotypes(LG = russianriver_lumpy_long_genos,
                                          CK = russianriver_ckmr_lumpy,
                                          max_mismatch = 6)
matchers
```

#### Check for full-sibs with bi-allelic data

```{r}
russianriver_FS_is <- mc_sample_simple(russianriver_Qs, 
                 nu = "FS",
                 de = "U")

russianriver_FS_is

#find where FNR > 0.01
russianriver_FS_is %>%
  filter(FNR == 0.01) %>%
  arrange(FNR) %>%
  slice(1)

```

Set lambda to 10.

```{r}
russianriver_lumpy_pairwise_logls <- pairwise_kin_logl_ratios(
                                              D1 = russianriver_lumpy_long_genos,
                                              D2 = russianriver_lumpy_long_genos,
                                              CK = russianriver_ckmr_lumpy,
                                              numer = "FS",
                                              denom = "U", 
                                              num_cores = 1)
```

Pairs with logLR greater than 10:

```{r}
russianriversibs<- russianriver_lumpy_pairwise_logls %>%
  filter(logl_ratio > 10) %>%
  arrange(desc(logl_ratio))
```

34 sibling pairs.

### Prep data for matrix

To use this data frame to construct a matrix, the whole genome data frame (converted to bi-allelic) needs to be pivoted wider so that individuals stay as rows, but the loci and allele copies become columns.

```{r}
russianriver_wider <- russianriver_lumpy_long_genos %>% 
  pivot_wider(
    names_from = c(Locus, gene_copy),
    values_from = Allele,
    names_sep = "."
  )
```

#### Allele dosages

To prepare it for AGH matrix, I now need to convert it into allele dosages. I wrote this code with the help of ChatGPT:

-   0 = zero copies of the reference allele

-   1 = one copy of the reference allele

-   2 = two copies of the reference allele

-   NA = missing genotype

```{r}
# Find the first allele-copy column for each locus, ending in .1
copy1_cols <- names(russianriver_wider)[str_ends(names(russianriver_wider), "\\.1$")]

# Start the dosage dataframe with individual ID (keep Indiv column)
russianriver_dosage <- data.frame(
  Indiv = russianriver_wider$Indiv
)

# Convert each pair of allele columns to dosage
for (col1 in copy1_cols) {
  
  # Find the matching second allele column
  col2 <- str_replace(col1, "\\.1$", ".2")
  
  # Get the locus name
  locus <- str_remove(col1, "\\.1$")
  
  # Find alleles present at this locus
  alleles <- sort(unique(c(
    russianriver_wider[[col1]],
    russianriver_wider[[col2]]
  )))
  
  # Remove NA from the list of alleles
  alleles <- alleles[!is.na(alleles)]
  
  # Check that the locus is biallelic
  if (length(alleles) > 2) {
    warning(
      paste(
        locus,
        "has", length(alleles),
        "alleles:", paste(alleles, collapse = ", ")
      )
    )
  }
  
  # Choose the first allele as the allele being counted
  ref_allele <- alleles[1]
  
  # Calculate dosage
  russianriver_dosage[[locus]] <- ifelse(
    is.na(russianriver_wider[[col1]]) |
      is.na(russianriver_wider[[col2]]),
    NA,
    (russianriver_wider[[col1]] == ref_allele) +
      (russianriver_wider[[col2]] == ref_allele)
  )
}

write.csv(russianriver_dosage, file = "Data/Populations/russianriver_alleledosage.csv")
```
